### 2025-03-13 paper/sections/experiments.qmd
Down to 16 pages without major cutting of content

1. To what extent does our proposed counterfactual training objective (@eq-obj) induce models to learn plausible explanations?
2. To what extent does our proposed counterfactual training objective (@eq-obj) yield more favorable algorithmic recourse outcomes in the presence of actionability constraints?
3. What are the effects of hyperparameter selection with respect to @eq-obj?

@@

We observe that for all datasets except *OL* and across all test settings, the average distance of counterfactuals from observed samples in the target class is reduced, indicating improved plausibility. The magnitude of improvements varies by dataset: for the simple synthetic datasets, distance reductions range from around 20-40% (*LS*, *Moon*) to almost 60% (*Circ*). For the real-world tabular datasets, improvements are generally smaller but still substantial in many cases with around 10-15% for *CH*, 11-28% for *GMSC*, 7-8% for *Cred* and around 3% for *Adult*. For our only vision dataset (*MNIST*), distances are reduced by up to 9%. The results for our proposed divergence metric are qualitatively similar, but generally even more pronounced: for the *Circ* dataset, implausibility is reduced by almost 94% to virtually zero as we verified by looking at the absolute outcome. Improvements for other datasets range from 28% (*Moon*) to 78% (*GMSC*). For *OL* the reduction is negative, consistent with the distance-based metric. The only dataset, for which our proposed metric disagrees with the distance-based metric is *MNIST*.

These broad and substantial improvements in plausibility generally do not come at the cost of decreased predictive performance: test accuracy for CT is virtually identical to the baseline for *Adult*, *Circ*, *LS*, *Moon* and *OL*, and even slightly improved for *Cred*. Exceptions to this general pattern are *MNIST*, *CH* and *GMSC*, for which we observe reduction in test accuracy of 2, 5 and 15 percentage points, respectively. We note in this context, that we have not optimized our models for predictive performance at all and worked with very small networks. In summary, we find that CT can substantially improve the quality of explanations learned by models without generally sacrificing predictive accuracy.

@@

### Impact of hyperparameter settings {#sec-hyperparameters}
We extensively test the impact of three types of hyperparameters on the proposed training regime. Our complete results are available in the technical appendix; this section focuses on the main findings.

@@

***Other hyperparameters.*** We observe that the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch. We also confirm that a higher number of training epochs is beneficial. Interestingly, we find that it is not necessary to employ CT during the entire training phase to achieve the desired improvements in explainability. When training models conventionally during the first 50% of epochs before switching to CT for the next 50% of epochs, we observed positive results. Put differently, CT may be a way to improve the explainability of models in a fine-tuning manner.

>>>>

1. To what extent does our proposed counterfactual training objective in @eq-obj induce models to learn plausible explanations?
2. To what extent does our proposed counterfactual training objective in @eq-obj yield more favorable algorithmic recourse outcomes in the presence of actionability constraints?
3. What are the effects of hyperparameter selection with respect to @eq-obj?

@@

We observe that for all datasets except *OL* and across all test settings, the average distance of counterfactuals from observed samples in the target class is reduced, indicating improved plausibility. The magnitude of improvements varies by dataset: for the simple synthetic datasets, distance reductions range from around 20-40% (*LS*, *Moon*) to almost 60% (*Circ*). For the real-world tabular datasets, improvements are generally smaller but still substantial in many cases with around 10-15% for *CH*, 11-28% for *GMSC*, 7-8% for *Cred* and around 3% for *Adult*. For our only vision dataset (*MNIST*), distances are reduced by up to 9%. The results for our proposed divergence metric are qualitatively similar, but generally even more pronounced: for the *Circ* dataset, implausibility is reduced by almost 94% to virtually zero as we verified by looking at the absolute outcome. Improvements for other datasets range from 28% (*Moon*) to 78% (*GMSC*). For *OL* the reduction is negative, consistent with the distance-based metric. *MNIST* is the only dataset for which the two metrics disagree.

These broad and substantial improvements in plausibility generally do not come at the cost of decreased predictive performance: test accuracy for CT is virtually identical to the baseline for *Adult*, *Circ*, *LS*, *Moon* and *OL*, and even slightly improved for *Cred*. Exceptions to this general pattern are *MNIST*, *CH* and *GMSC*, for which we observe reduction in test accuracy of 2, 5 and 15 percentage points, respectively. We note in this context, that we have not optimized our models for predictive performance at all and worked with very small networks. In summary, we find that CT can substantially improve the quality of explanations learned by models without sacrificing predictive accuracy.

@@

### Impact of hyperparameter settings {#sec-hyperparameters}
We test in-depth the impact of three types of hyperparameters; our complete results are in the appendix.

@@

***Other hyperparameters.*** We observe that the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch. We also confirm that a higher number of training epochs is beneficial. Interestingly, we observed desired improvements in explainability when CT was combined with conventional training and applied only for the final 50% of epochs of the complete training process. Put differently, CT may be a way to improve the explainability of models in a fine-tuning manner.

### 2025-03-13 paper/sections/intro.qmd
Down to 16 pages without major cutting of content

The trend towards complexity for the sake of performance has come under serious scrutiny in recent years. At the very cusp of the deep learning revolution, @szegedy2013intriguing showed that artificial neural networks (ANN) are sensitive to adversarial examples: perturbed versions of data instances that yield vastly different model predictions despite being "imperceptible" in that they are semantically indifferent from their factual counterparts. Even though some partially effective mitigation strategies have been proposed---most notably **adversarial training** [@goodfellow2014explaining]---truly robust deep learning (DL) remains unattainable even for models that are considered shallow by today's standards [@kolter2023keynote].

Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] Indeed, recent work on the so-called "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% while preserving their predictive performance [@frankle2018lottery] and generalizability [@morcos2019success]. Similarly, @zhang2021understanding showed that state-of-the-art neural networks are so expressive that they can fit randomly labeled data. Thus, looking at the predictive performance alone, the solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns. This poses two related challenges. Firstly, there is no dependable way to verify if such complex representations correspond to meaningful and plausible explanations. Secondly, even if we could resolve the first challenge, it remains undecided how to ensure that models can *only* learn valuable explanations.

@@

**Specifically, we introduce counterfactual training (CT)**: a novel training regime explicitly meant to align learned representations with plausible explanations that comply with user requirements. The remainder of this paper is structured as follows. @sec-lit presents related work, focusing in particular on the link between adversarial examples and counterfactual explanations. Then follow our main contributions:

1. In @sec-method, we introduce our methodological framework and show theoretically that it can be used to enforce global actionability constraints.  
2. Through extensive experiments we demonstrate that CT substantially improves explainability without sacrificing predictive performance (@sec-experiments).

We discuss future research and challenges in @sec-discussion and conclude in @sec-conclusion that CT is a promising new approach towards making opaque models more trustworthy.

>>>>

The trend towards complexity for the sake of performance has come under serious scrutiny in recent years. At the very cusp of the deep learning (DL) revolution, @szegedy2013intriguing showed that artificial neural networks (ANN) are sensitive to adversarial examples (AEs): perturbed versions of data instances that yield vastly different model predictions despite being "imperceptible" in that they are semantically indifferent from their factual counterparts. Even though some partially effective mitigation strategies have been proposed---most notably **adversarial training** [@goodfellow2014explaining]---truly robust deep learning remains unattainable even for models that are considered "shallow" by today's standards [@kolter2023keynote].

Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] Indeed, recent work on the so-called "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% while preserving their predictive performance [@frankle2018lottery]. Similarly, @zhang2021understanding showed that state-of-the-art neural networks are expressive enough to fit randomly labeled data. Thus, looking at the predictive performance alone, the solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns. This poses two challenges. Firstly, there is no dependable way to verify if representations correspond to meaningful, plausible explanations. Secondly, even if we could resolve the first challenge, it remains undecided how to ensure that models can *only* learn valuable explanations.

@@

**Specifically, we propose counterfactual training (CT)**: a novel training regime that aligns learned representations with plausible explanations compliant with user requirements. The remainder of this paper is structured as follows. @sec-lit presents related work, focusing on the link between adversarial examples and counterfactual explanations. Then follow our main contributions:

1. In @sec-method, we introduce our methodological framework and show theoretically that it can be used to enforce global actionability constraints.  
2. In @sec-experiments, through extensive experiments we demonstrate that CT substantially improves explainability without sacrificing predictive performance.

We discuss the challenges in @sec-discussion and conclude in @sec-conclusion that CT is a promising approach towards making opaque models more trustworthy.

### 2025-03-13 paper/sections/method.qmd
Down to 16 pages without major cutting of content

$$
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
$$ {#eq-obj}
where $\text{yloss}(\cdot)$ is a classification loss that induces discriminative performance (e.g., cross-entropy). The second and third terms in @eq-obj are explained in detail below. For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and, (2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit]. The trade-off between the components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.

@@

It is worth highlighting that Prp.~\ref{prp-mtblty} assumes independence of features. This raises a valid concern about the effect of protecting immutable features in the presence of proxies that remain unprotected. We address this in @sec-discussion.

>>>>

$$
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
$$ {#eq-obj}
where $\text{yloss}(\cdot)$ is a classification loss that induces discriminative performance (e.g., cross-entropy). The second and third terms are explained in detail below. For now, they can be summarized as inducing explainability directly and indirectly by penalizing: (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and, (2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit]. The trade-off between the components can governed through penalties $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.

@@

It is worth highlighting that \mbox{Prp. \ref{prp-mtblty}} assumes independence of features. This raises a valid concern about the effect of protecting immutable features in the presence of proxies that remain unprotected. We address this in @sec-discussion.

### 2025-03-13 paper/sections/related.qmd
Down to 16 pages without major cutting of content

To the best of our knowledge, the proposed framework for counterfactual training represents the first attempt to use counterfactual explanations during training to improve model explainability. In high-level terms, we define model explainability as the extent to which valid explanations derived for an opaque model are also deemed plausible with respect to the underlying data and stakeholder requirements; the former means that the counterfactuals should comply with the distribution of the factual data, the latter means that they should respect arbitrary (global) actionability constraints. To make the desiderata for our framework more concrete, we follow @augustin2020adversarial in tying the concept of explainability to the quality of counterfactual explanations that we can generate for a given model. The authors show that CEs---understood here as minimal input perturbations that yield some desired model prediction---are generally more meaningful if the underlying model is more robust to adversarial examples. We can make intuitive sense of this finding when looking at adversarial training (AT) through the lens of representation learning with high degrees of freedom. As argued before, learned representations may be sensitive to producing implausible explanations and mispredicting for worst-case counterfactuals (i.e., adversarial examples). Thus, by inducing models to "unlearn" susceptiblity to such examples, AT can effectively remove implausible explanations from the solution space.

@@

There have also been several attempts at formalizing the relationship between counterfactual explanations and adversarial examples (AE). Pointing to clear similarities in how CEs and AEs are generated, @freiesleben2022intriguing makes the case for jointly studying the opaqueness and robustness problems in representation learning. Formally, AEs can be seen as the subset of CEs for which misclassification is achieved [@freiesleben2022intriguing]. Similarly, @pawelczyk2022exploring show that CEs and AEs are equivalent under certain conditions and derive theoretical upper bounds on distances between them.

>>>>

To the best of our knowledge, the proposed framework for counterfactual training represents the first attempt to use counterfactual explanations during training to improve model explainability. In high-level terms, we define model explainability as the extent to which valid explanations derived for an opaque model are also deemed plausible with respect to the underlying data and (global) actionability constraints. To make the desiderata for our framework more concrete, we follow @augustin2020adversarial in tying the concept of explainability to the quality of CEs that can be generated for a given model. The authors show that CEs---understood as minimal input perturbations that yield some desired model prediction---tend to be more meaningful if the underlying model is more robust to adversarial examples. We can make intuitive sense of this finding when looking at adversarial training (AT) through the lens of representation learning with high degrees of freedom. As argued before, learned representations may be sensitive to producing implausible explanations and mispredicting for worst-case counterfactuals (i.e., AEs). Thus, by inducing models to "unlearn" susceptiblity to such examples, AT can effectively remove implausible explanations from the solution space.

@@

There have also been several attempts at formalizing the relationship between counterfactual explanations and adversarial examples. Pointing to clear similarities in how CEs and AEs are generated, @freiesleben2022intriguing makes the case for jointly studying the opaqueness and robustness problems in representation learning. Formally, AEs can be seen as the subset of CEs for which misclassification is achieved [@freiesleben2022intriguing]. Similarly, @pawelczyk2022exploring show that CEs and AEs are equivalent under certain conditions and derive theoretical upper bounds on distances between them.
